---
title: "Cantaloupe_git"
author: "Sophia"
Date: 12.11.2017
output:
  html_document:
    code_folding: hide
    toc: true
---
```{r echo=F, message=F, warning=F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(ggthemes)
```


#Experimental Design 
 * One cantaloupe rind was inoculated with 100uL bacterial culture
 * 1 single rind was placed in one petri dish and sealed and processed in falcon tube
 * 10mL of PBS was added into each falcon tube for processing
 * Dilutions of wash water was plated


# Import and clean up the spreadsheet
 *All Variables need to be factor or numeric
 *Substitute detection limits, lower: 100 CFU, upper: 4e5
 *Raw count in excel is CFU/mL from Qcount, taken into account dilutions but not 10mL PBS
 
```{r}
cantaloupe <- read.csv("Cantaloupe_Summary.csv")
names(cantaloupe)[1] <- "Condition"
#make sure all variables are factor or numeric if needed
cantaloupe_adj <- cantaloupe %>%
  mutate(Replicate=as.factor(Replicate), Dilution=as.numeric(Dilution))  %>%
# Substitute upper detection limit for tntc counts
  mutate(Raw=ifelse(tnt.cnt.no_growth=="tnt",(4e5)*10^Dilution,Raw)) %>%
  # Substitute lower detection limit for zero counts
  mutate(Raw=ifelse(tnt.cnt.no_growth=="no_growth",10*10^Dilution,Raw)) %>%
  # Calculate inoculum per-rind: Raw/10, since 100uL was added per rind
  # Calculate wash water per-rind: Raw*10
  mutate(cfu_cantaloupe=ifelse(Day=="Inoculum",Raw/1000,Raw*10)) %>%
    filter(!(Condition=="Stationary" & Replicate=="2" & Day=="Inoculum"))  


#proper labeling of strains and setting order
cantaloupe_adj$Day2 <- factor(cantaloupe_adj$Day, levels=c("Inoculum", "0", "1", "3", "4", "7")) #re-order how it is displayed
levels(cantaloupe_adj$Strain)[levels(cantaloupe_adj$Strain)=="Inoculum"] <- "Inoc"
levels(cantaloupe_adj$Strain)[levels(cantaloupe_adj$Strain)=="FSL C2-0008"] <- "FSL C2-0008*"
cantaloupe_adj$Strain2 <- factor(cantaloupe_adj$Strain, levels=c("FSL J1-031", "FSL R9-0506", "FSL R9-5411", "FSL R9-5506", "FSL C2-0008*", "APC")) #re-order how it is displayed


cantaloupe_adj %>% 
   filter(Strain != "LMPM") %>%
  filter(Strain != "MOX") %>%
  select(Strain2, Condition, Day2, cfu_cantaloupe)%>%
  group_by(Strain2, Condition, Day2) %>%
  filter(!is.na(cfu_cantaloupe)) -> cantaloupe_adj_2

cantaloupe_adj_2<-filter(cantaloupe_adj_2, cfu_cantaloupe!=0)


cantaloupe_adj_2 %>%
  summarize(average=mean(log10(cfu_cantaloupe)),
            SDE=sd(log10(cfu_cantaloupe))/sqrt(n()),
            n=n(),
            zeroes=sum(cfu_cantaloupe==0)) -> cantaloupeSummary

```


#Arithemtic mean for aerobic plate count (APC)
```{r}
cantaloupe_adj %>%  
   filter(Strain=="APC") %>% 
    filter(Day !="2")%>% #one sample was processed on day 2
    filter(!is.na(cfu_cantaloupe)) -> cantaloupeSummary_APC

cantaloupeSummary_APC<-filter(cantaloupeSummary_APC, cfu_cantaloupe!=0)

cantaloupeSummary_APC %>%
group_by(Day) %>%
  summarize(average=mean(log10(cfu_cantaloupe)),
            SDE=sd(log10(cfu_cantaloupe))/sqrt(n()),
            n=n(),
            zeroes=sum(cfu_cantaloupe==0)) -> cantaloupeSummary_APC_2

```

#Plots of raw counts
```{r}
#Plot Strain~Condition with each of the three replicates
cantaloupe_adj %>%
  filter(Strain != "LMPM") %>%
  filter(Strain != "MOX") %>%
  mutate(Day_n = as.numeric(as.character(Day))) %>%
  ggplot(aes(x=Day_n,y=cfu_cantaloupe,color=Replicate)) +
  geom_point() +
  geom_line() +
  scale_y_log10() +
  facet_grid(Strain~Condition)


#Plot with arithemtic mean for Strain~Condition and add fancy ribbon for standard deviation, write in bold processing days
cantaloupeSummary %>%
#filter(Strain != "LMPM") %>%
 # filter(Strain != "MOX") %>%
 # filter(Day2 != "2") %>%
  ggplot(aes(x=Day2,y=average, color=Strain2, group=1)) +
 guides(color=FALSE)+
    labs(x="Day",y="log CFU/ cantaloupe rind",title=expression(italic("Listeria")))+
  geom_point() +
    geom_ribbon(aes(ymin=average-SDE,
                  ymax=average+SDE,fill=Strain2, linetype=NA), alpha=0.2) + 
   scale_x_discrete(limits=c("Inoc","0","1","2","3","4","5", "6","7"))+
  theme(axis.text.x = element_text(face = c('bold', 'bold', 'bold', 'plain', 'bold', 'bold', 'plain', 'plain', 'bold'), size = c('10.8', '11', '11', '9', '11', '11', '9', '9', '11'), margin=margin(-35,0,0,0)),legend.position="none")+
  # labs( caption="Ribbon: Standard deviation")+
     labs( caption="* Surrogate organism")+
  facet_grid(Condition~Strain2) 
ggsave("Cantaloupe.png", width=18, height=9,  dpi=600, units=c("in"))
```


#Population dynamics will be assessed for three phases (i) initial die-off, (ii) change in bacterial numbers over the initial 24 h, and (iii) growth or die-off over 7 day storage

# (i) Initial die-off
```{r}
cantaloupe_dieoff_sur <- cantaloupe_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_cantaloupe) %>%
  filter(Strain!="APC") %>%
 filter(Strain != "LMPM") %>%
  filter(Strain != "MOX") %>%
   filter(!(Condition=="Stationary" & Replicate=="2" & Day=="Inoculum")) %>% #invalid
  spread(key=Day,value=cfu_cantaloupe,sep="_") %>%
  mutate(first_dieoff_sur=log10(Day_0/Day_Inoculum))


# Fit (fixed-effect) linear model
m_dieoff_anova_sur <- lm(first_dieoff_sur~Strain*Condition,
                  data=cantaloupe_dieoff_sur)
anova(m_dieoff_anova_sur)
#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_dieoff_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1

# Since interaction is not significant, remove before doing pairwise tests:
m_dieoff_anova_sur <- update(m_dieoff_anova_sur,.~.-Strain:Condition)
anova(m_dieoff_anova_sur)
qqnorm(resid(m_dieoff_anova_sur))
qqline(resid(m_dieoff_anova_sur))
plot(predict(m_dieoff_anova_sur),resid(m_dieoff_anova_sur))

first_dieoff_lsm_condition <- lsmeans(m_dieoff_anova_sur,~Condition)
first_dieoff_pairwise_condition <- lsmeans(first_dieoff_lsm_condition,pairwise~Condition)
condition_cld_dieoff <- cld(first_dieoff_pairwise_condition[[1]], Letters=LETTERS) %>% select(Condition,.group)
summary(first_dieoff_lsm_condition) %>%
  data.frame() %>%
  left_join(condition_cld_dieoff,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) + labs(x="Condition",y="estimated means log reduction",title=(expression("Initial die-off "(italic("Listeria"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#and now by strain
first_dieoff_lsm_strain <- lsmeans(m_dieoff_anova_sur,~Strain)
first_dieoff_pairwise_strain <- lsmeans(first_dieoff_lsm_strain,pairwise~Strain)
strain_cld_dieoff <- cld(first_dieoff_pairwise_strain[[1]], Letters=LETTERS) %>% select(Strain,.group)
summary(first_dieoff_lsm_strain) %>%
  data.frame() %>%
  left_join(strain_cld_dieoff,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)


#Estimated means log CFU/lettuce with confidence intervals for supplemental table 3
summary(first_dieoff_lsm_strain)
summary(first_dieoff_lsm_condition)
```

# (ii) change in bacterial numbers over the initial 24 h
```{r}

cantaloupe_24h_sur <- cantaloupe_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_cantaloupe) %>%
  filter(Strain!="APC") %>%
 filter(Strain != "LMPM") %>%
  filter(Strain != "MOX") %>%
  spread(key=Day,value=cfu_cantaloupe,sep="_") %>%
  mutate(first_24h_sur=log10(Day_1/Day_0))

cantaloupe_24h_sur %>%
  ggplot(aes(x=Condition,y=first_24h_sur,color=Strain)) +
  geom_point(position=position_dodge(width=.5))


# Fit (fixed-effect) linear model
m_24h_anova_sur <- lm(first_24h_sur~Strain*Condition,
                  data=cantaloupe_24h_sur)
anova(m_24h_anova_sur)

#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_24h_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1

# Since interaction is not significant, remove before doing pairwise tests:
m_24h_anova_sur <- update(m_24h_anova_sur,.~.-Strain:Condition)
anova(m_24h_anova_sur)
qqnorm(resid(m_24h_anova_sur))
qqline(resid(m_24h_anova_sur))
plot(predict(m_24h_anova_sur),resid(m_24h_anova_sur))


first_24h_lsm_condition <- lsmeans(m_24h_anova_sur,~Condition)
first_24h_pairwise_condition <- lsmeans(first_24h_lsm_condition,pairwise~Condition)
condition_cld_24h <- cld(first_24h_pairwise_condition[[1]], Letters=LETTERS) %>% select(Condition,.group)
summary(first_24h_lsm_condition) %>%
  data.frame() %>%
  left_join(condition_cld_24h,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) + labs(x="Condition",y="estimated means log reduction",title=(expression("First 24 hr "(italic("Listeria"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#and now by strain
first_24h_lsm_strain <- lsmeans(m_24h_anova_sur,~Strain)
first_24h_pairwise_strain <- lsmeans(first_24h_lsm_strain,pairwise~Strain)
strain_cld_24h <- cld(first_24h_pairwise_strain[[1]], Letters=LETTERS) %>% select(Strain,.group)
summary(first_24h_lsm_strain) %>%
  data.frame() %>%
  left_join(strain_cld_24h,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/lettuce with confidence intervals for supplemental table 3
summary(first_24h_lsm_strain)
summary(first_24h_lsm_condition)

```

(iii) growth over 7 day storage
#N7-Nmin
```{r}
cantaloupe_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_cantaloupe) %>%
  filter(Strain!="APC") %>%
   filter(Strain != "LMPM") %>%
  filter(Strain != "MOX") %>%
  spread(key=Day,value=cfu_cantaloupe,sep="_") -> cantaloupe_adj_day

cantaloupe_Nmin <- transform(cantaloupe_adj_day, Nmin=pmin(Day_0, Day_1, Day_3, Day_4, Day_7))
cantaloupe_Nmin %>%
  mutate(compare_Day7_Nmin=log10(Day_7/Nmin)) -> cantaloupe_N7Nmin_sur

#Average increase
cantaloupe_N7Nmin_sur %>%
  filter(!(compare_Day7_Nmin=="NA")) %>%
  summarize(average_increase=mean(compare_Day7_Nmin),  n=n()) -> compare_Day7_Nmin_cant



# Fit (fixed-effect) linear model
m_N7Nmin_anova_sur <- lm(compare_Day7_Nmin~Strain*Condition,
                  data=cantaloupe_N7Nmin_sur)
anova(m_N7Nmin_anova_sur)
#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_N7Nmin_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1

# Since interaction is not significant, remove before doing pairwise tests:
m_N7Nmin_anova_sur <- update(m_N7Nmin_anova_sur,.~.-Strain:Condition)
anova(m_N7Nmin_anova_sur)
qqnorm(resid(m_N7Nmin_anova_sur))
qqline(resid(m_N7Nmin_anova_sur))
plot(predict(m_N7Nmin_anova_sur),resid(m_N7Nmin_anova_sur))


first_N7Nmin_lsm_condition <- lsmeans(m_N7Nmin_anova_sur,~Condition)
first_N7Nmin_pairwise_condition <- lsmeans(first_N7Nmin_lsm_condition,pairwise~Condition)
condition_cld_N7Nmin <- cld(first_N7Nmin_pairwise_condition[[1]], Letters=LETTERS) %>% select(Condition,.group)
summary(first_N7Nmin_lsm_condition) %>%
  data.frame() %>%
  left_join(condition_cld_N7Nmin,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) + labs(x="Condition",y="estimated means log reduction",title=(expression("Re-growth to day 7 "(italic("Listeria"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#and now by strain
first_N7Nmin_lsm_strain <- lsmeans(m_N7Nmin_anova_sur,~Strain)
first_N7Nmin_pairwise_strain <- lsmeans(first_N7Nmin_lsm_strain,pairwise~Strain)
strain_cld_N7Nmin <- cld(first_N7Nmin_pairwise_strain[[1]], Letters=LETTERS) %>% select(Strain,.group)
summary(first_N7Nmin_lsm_strain) %>%
  data.frame() %>%
  left_join(strain_cld_N7Nmin,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/lettuce with confidence intervals for supplemental table 3
summary(first_N7Nmin_lsm_strain)
summary(first_N7Nmin_lsm_condition)

```


#We want to determine Days for Daymin and Daymax

```{r}
cantaloupe_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_cantaloupe) %>%
  filter(Strain!="APC") %>%
   filter(Strain != "LMPM") %>%
  filter(Strain != "MOX") %>%
  filter(Day != "Inoculum") %>%
  filter(Day != "0") %>%
  filter(Day !="2")%>%
  spread(key=Day,value=cfu_cantaloupe,sep="_") -> cantaloupe_adj_Days

#----min values
cantaloupe_adj_Days_Nmin <- transform(cantaloupe_adj_Days, Nmin=pmin(Day_1, Day_3, Day_4, Day_7))

cantaloupe_adj_Days_Nmin->a_min

a_min$CompareColumn <- apply(a_min[, 4:7] == a_min[, 8], 1, function(x) ifelse(any(x), paste(colnames(a_min)[4:7][x], collapse=', '), 'None'))

count(a_min, c("CompareColumn")) ->b_min

#----- max value
cantaloupe_adj_Days_Nmax <- transform(cantaloupe_adj_Days, Nmax=pmax(Day_1, Day_3, Day_4, Day_7))

cantaloupe_adj_Days_Nmax->a_max

a_max$CompareColumn <- apply(a_max[, 4:7] == a_max[, 8], 1, function(x) ifelse(any(x), paste(colnames(a_max)[4:7][x], collapse=', '), 'None'))

count(a_max, c("CompareColumn")) ->b_max
```

# Net growth
```{r}
cantaloupe_adj%>%
  select(Strain,Condition,Replicate,Day,cfu_cantaloupe) %>%
 filter(Strain!="APC") %>%
     filter(Strain!="LMPM") %>%
   filter(Strain!="MOX") %>%
    filter(!is.na(cfu_cantaloupe)) %>%
  spread(key=Day,value=cfu_cantaloupe,sep="_") %>%
  mutate(netgrowth=log10(Day_7/Day_Inoculum)) -> cantaloupe_adj2



cantaloupe_adj2 %>%
  filter(!is.na(netgrowth))%>%
  group_by(Strain, Condition) %>%
  summarize(average=mean((netgrowth)),
            SDE=sd((netgrowth))/sqrt(n()),
            n=n()) -> cantaloupeSummary_netgrowth

format.data.frame(cantaloupeSummary_netgrowth, dec=".") ->netgrowth_Listeria
write_xlsx(netgrowth_Listeria, "netgrowth_Listeria.xlsx")
```
