---
title: "lettuce_git"
author: "Sophia"
Date: 01.08.2017
output:
  html_document:
    code_folding: hide
    toc: true
---


```{r echo=F, message=F, warning=F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(lsmeans)
library(ggthemes)
```

#Experimental Design 
 * 3-3.5g lettuce were inoculated with 100uL bacterial culture
 * Lettuce was placed in a petri-dish
 * 47mL of PBS was added into each bag for processing
 * Dilutions of wash water was plated


# Import and clean up the spreadsheet
 *All Variables need to be factor or numeric
 *Substitute detection limits, lower: 470CFU, upper: 4e5
 
 
 
```{r}
lettuce <- read.csv("Lettuce_Summary_mod.csv") %>%
  filter(Strain!="MacConkey") 
 # filter(Strain!="APC")
names(lettuce)[1] <- "Condition"
#make sure all variables are factor or numeric if needed
lettuce_adj <- lettuce %>%
  mutate(Replicate=as.factor(Replicate), Dilution=as.numeric(Dilution))  %>%
  # Substitute upper detection limit for tntc counts
  mutate(Raw=ifelse(tnt.cnt.no.growth=="tntc",(4e5)*10^Dilution,Raw)) %>%
  # Substitute lower detection limit for zero counts
  mutate(Raw=ifelse(Raw==0,1/(v*2)*10^Dilution,Raw)) %>%
  # Calculate inoculum per-lettuce: Raw/10, since 100uL was added per lettuce
  # Calculate wash water per-lettuce: Raw*47
  mutate(cfu_lettuce=ifelse(Day=="Inoculum",Raw/100,Raw*47)) 


lettuce_adj %>%  
  group_by(Strain, Condition, Day) %>%
  filter(!is.na(cfu_lettuce)) %>%
  summarize(average=mean(log10(cfu_lettuce)),
            SDE=sd(log10(cfu_lettuce))/sqrt(n()),
            n=n(),
            zeroes=sum(cfu_lettuce==0)) -> lettuceSummary
```

#Arithemtic mean for aerobic plate count (APC)
```{r}
lettuce_APC <- lettuce_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_lettuce) %>%
  filter(Strain=="APC") %>%
  spread(key=Day,value=cfu_lettuce,sep="_") %>%
  mutate(APC_day7_day0=log10(Day_7/Day_0))

View(lettuce_APC)

lettuce_APC %>%  
  summarize(average=mean(APC_day7_day0),
            SDE=sd(APC_day7_day0)/sqrt(n()),
            n=n()) -> lettuce_APC_adj
View(lettuce_APC_adj)
```

#Plots of raw counts
```{r}
#Plot Strain~Condition with each of the three replicates
lettuce_adj %>%
  filter(Day != "Inoculum") %>%
  filter(Strain != "MacConkey") %>%
  mutate(Day_n = as.numeric(as.character(Day))) %>%
  ggplot(aes(x=Day_n,y=cfu_lettuce,color=Replicate)) +
  geom_point() +
  geom_line() +
  scale_y_log10() +
  facet_grid(Strain~Condition)

#Plot with arithemtic mean for Strain~Condition and add fancy ribbon for standard deviation, write in bold processing days

#Write out full Food Safety Lab (FSL) strain ID
lettuceSummary$Strain2 <- factor(lettuceSummary$Strain, levels=c("FSL R9-5258", "FSL R9-5271", "FSL R9-5515", "FSL R9-5516", "FSL R9-5517","FSL R9-4077", "APC")) #re-order how it is displayed
lettuceSummary$Day2 <- factor(lettuceSummary$Day, levels=c("Inoculum", "0", "1", "3", "4", "7")) #re-order how it is displayed
levels(lettuceSummary$Day2)[levels(lettuceSummary$Day2)=="Inoculum"] <- "Inoc"
levels(lettuceSummary$Strain2)[levels(lettuceSummary$Strain2)=="FSL R9-4077"] <- "FSL R9-4077*"

lettuceSummary %>%
 filter(Strain != "MacConkey") %>%
  ggplot(aes(x=Day2,y=average, color=Strain2, group=1)) +
 guides(color=FALSE)+
    labs(x="Day",y="log CFU/lettuce",title=expression(italic("E. coli"))) +
  geom_point() +
    geom_ribbon(aes(ymin=average-SDE,
                  ymax=average+SDE,fill=Strain2, linetype=NA), alpha=0.2) + 
   scale_x_discrete(limits=c("Inoc","0","1","2","3","4","5", "6","7"))+
  theme(axis.text.x = element_text(face = c('bold', 'bold', 'bold', 'plain', 'bold', 'bold', 'plain', 'plain', 'bold'), size = c('10.8', '11', '11', '9', '11', '11', '9', '9', '11'),margin=margin(-35,0,0,0)),legend.position="none")+
  # labs( caption="Ribbon: Standard deviation")+
   labs( caption="* Surrogate Organism")+
  facet_grid(Condition~Strain2) 
ggsave("Lettuce.png", width=18, height=9,  dpi=600, units=c("in"))
```

#Population dynamics will be assessed for three phases (i) initial die-off, (ii) change in bacterial numbers over the initial 24 h, and (iii) growth or die-off over 7 day storage

# (i) Initial die-off
```{r}

lettuce_die_off_sur <- lettuce_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_lettuce) %>%
    filter(Strain!="APC") %>%
    filter(Strain!="MacConkey") %>%
   filter(!(Strain=="FSL R9-5516" & Condition=="Glyc" & Replicate=="1")) %>% #inoculum culture didn't grow
   filter(!(Strain=="FSL R9-5271" & Condition=="NaCl" & Replicate=="2" & Day=="Inoculum")) %>% #invalid
  filter(!(Strain=="FSL R9-4077" & Condition=="NaCl" & Replicate=="3" & Day=="Inoculum")) %>% #invalid
  spread(key=Day,value=cfu_lettuce,sep="_") %>%
  mutate(initial_die_off_sur=log10(Day_0/Day_Inoculum))


# Fit (fixed-effect) linear model
m_die_off_anova_sur <- lm(initial_die_off_sur~Strain*Condition,
                  data=lettuce_die_off_sur)
anova(m_die_off_anova_sur)
#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_die_off_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1


# Since interaction is not significant, remove before doing pairwise tests:
m_die_off_anova_sur <- update(m_die_off_anova_sur,.~.-Strain:Condition)
anova(m_die_off_anova_sur)
qqnorm(resid(m_die_off_anova_sur))
qqline(resid(m_die_off_anova_sur))
plot(predict(m_die_off_anova_sur),resid(m_die_off_anova_sur))

die_off_lsm_condition_sur <- lsmeans(m_die_off_anova_sur,~Condition)
die_off_pairwise_condition_sur <- lsmeans(die_off_lsm_condition_sur,pairwise~Condition)
condition_cld <- cld(die_off_pairwise_condition_sur[[1]],Letters=LETTERS) %>% select(Condition,.group)
summary(die_off_lsm_condition_sur) %>%
  data.frame() %>%
  left_join(condition_cld,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) +  labs(x="Condition",y="estimated means log reduction",title=(expression("Initial die-off "(italic("E. coli"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#by strain
die_off_lsm_strain_sur <- lsmeans(m_die_off_anova_sur,~Strain)
die_off_pairwise_strain_sur <- lsmeans(die_off_lsm_strain_sur,pairwise~Strain)
strain_cld <- cld(die_off_pairwise_strain_sur[[1]],Letters=LETTERS) %>% select(Strain,.group)

summary(die_off_lsm_strain_sur) %>%
  data.frame() %>%
  left_join(strain_cld,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/lettuce with confidence intervals for supplemental table 2
summary(die_off_lsm_strain_sur)
summary(die_off_lsm_condition_sur)

```

# (ii) change in bacterial numbers over the initial 24 h
```{r}
lettuce_24h_sur <- lettuce_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_lettuce) %>%
  filter(Strain!="APC") %>%
  filter(Strain!="MacConkey") %>%
  filter(!(Strain=="FSL R9-5516" & Condition=="Glyc" & Replicate=="1")) %>% #inoculum culture didn't grow
  spread(key=Day,value=cfu_lettuce,sep="_") %>%
  mutate(first_24h_sur=log10(Day_1/Day_0))

# Fit (fixed-effect) linear model
m_24h_anova_sur <- lm(first_24h_sur~Strain*Condition,
                  data=lettuce_24h_sur)
anova(m_24h_anova_sur)
#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_24h_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1

# Since interaction is not significant, remove before doing pairwise tests:
m_24h_anova_sur <- update(m_24h_anova_sur,.~.-Strain:Condition)
anova(m_24h_anova_sur)
qqnorm(resid(m_24h_anova_sur))
qqline(resid(m_24h_anova_sur))
plot(predict(m_24h_anova_sur),resid(m_24h_anova_sur))

first_24h_lsm_condition_sur <- lsmeans(m_24h_anova_sur,~Condition)
first_24h_pairwise_condition_sur <- lsmeans(first_24h_lsm_condition_sur,pairwise~Condition)
condition_cld_24h_sur <- cld(first_24h_pairwise_condition_sur[[1]],Letters=LETTERS) %>% select(Condition,.group)
summary(first_24h_lsm_condition_sur) %>%
  data.frame() %>%
  left_join(condition_cld_24h_sur,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) + labs(x="Condition",y="estimated means log CFU/lettuce",title=(expression("First 24 hr "(italic("E. coli"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#now by strain
first_24h_lsm_strain_sur <- lsmeans(m_24h_anova_sur,~Strain)
first_24h_pairwise_strain_sur <- lsmeans(first_24h_lsm_strain_sur,pairwise~Strain)
strain_cld_24h_sur <- cld(first_24h_pairwise_strain_sur[[1]],Letters=LETTERS) %>% select(Strain,.group)
summary(first_24h_lsm_strain_sur) %>%
  data.frame() %>%
  left_join(strain_cld_24h_sur,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) + 
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/lettuce with confidence intervals for supplemental table 2
summary(first_24h_lsm_strain_sur)
summary(first_24h_lsm_condition_sur)
```

(iii) die-off over 7 day storage
```{r}
lettuce_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_lettuce) %>%
  filter(Strain!="APC") %>%
  filter(Strain!="MacConkey") %>%
    filter(!(Strain=="FSL R9-5516" & Condition=="Glyc" & Replicate=="1")) %>% #inoculum didn't grow
  spread(key=Day,value=cfu_lettuce,sep="_") %>%
  mutate(Nday0_Nday7=log10(Day_7/Day_0)) -> Nday0Nday7_lettuce

# Fit (fixed-effect) linear model
m_Nday0_Nday7_anova_sur <- lm(Nday0_Nday7~Strain*Condition,
                  data=Nday0Nday7_lettuce)
anova(m_Nday0_Nday7_anova_sur)
#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_Nday0_Nday7_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1

# Since interaction is not significant, remove before doing pairwise tests:
m_Nday0_Nday7_anova_sur <- update(m_Nday0_Nday7_anova_sur,.~.-Strain:Condition)
anova(m_Nday0_Nday7_anova_sur)
qqnorm(resid(m_Nday0_Nday7_anova_sur))
qqline(resid(m_Nday0_Nday7_anova_sur))
plot(predict(m_Nday0_Nday7_anova_sur),resid(m_Nday0_Nday7_anova_sur))

Nday0_Nday7_lsm_condition_sur <- lsmeans(m_Nday0_Nday7_anova_sur,~Condition)
Nday0_Nday7_pairwise_condition_sur <- lsmeans(Nday0_Nday7_lsm_condition_sur,pairwise~Condition)
Nday0_Nday7_cld <- cld(Nday0_Nday7_pairwise_condition_sur[[1]],Letters=LETTERS) %>% select(Condition,.group)
summary(Nday0_Nday7_lsm_condition_sur) %>%
  data.frame() %>%
  left_join(Nday0_Nday7_cld,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) +  labs(x="Condition",y="estimated means log CFU/lettuce",title=(expression("Nday7-Nday0 "(italic("E. coli"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)


#and now by strain
Nday0_Nday7_lsm_strain_sur <- lsmeans(m_Nday0_Nday7_anova_sur,~Strain)
Nday0_Nday7_pairwise_strain_sur <- lsmeans(Nday0_Nday7_lsm_strain_sur,pairwise~Strain)
Nday0_Nday7_strain_cld <- cld(Nday0_Nday7_pairwise_strain_sur[[1]],Letters=LETTERS) %>% select(Strain,.group)

summary(Nday0_Nday7_lsm_strain_sur) %>%
  data.frame() %>%
  left_join(Nday0_Nday7_strain_cld,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/lettuce with confidence intervals for supplemental table 2
summary(Nday0_Nday7_lsm_condition_sur)
summary(Nday0_Nday7_lsm_strain_sur)


```

#We want to determine Days for Daymin and Daymax

```{r}
lettuce_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_lettuce) %>%
  filter(Strain!="APC") %>%
    filter(Strain!="MacConkey") %>%
  filter(Day != "Inoculum") %>%
  filter(Day != "0") %>%
  spread(key=Day,value=cfu_lettuce,sep="_") -> lettuce_adj_Days

#----min values
lettuce_adj_Days_Nmin <- transform(lettuce_adj_Days, Nmin=pmin(Day_1, Day_3, Day_4, Day_7))

lettuce_adj_Days_Nmin->a_min

a_min$CompareColumn <- apply(a_min[, 4:7] == a_min[, 8], 1, function(x) ifelse(any(x), paste(colnames(a_min)[4:7][x], collapse=', '), 'None'))

count(a_min, c("CompareColumn")) ->b_min

#----- max value
lettuce_adj_Days_Nmax <- transform(lettuce_adj_Days, Nmax=pmax(Day_1, Day_3, Day_4, Day_7))

lettuce_adj_Days_Nmax->a_max

a_max$CompareColumn <- apply(a_max[, 4:7] == a_max[, 8], 1, function(x) ifelse(any(x), paste(colnames(a_max)[4:7][x], collapse=', '), 'None'))

count(a_max, c("CompareColumn")) ->b_max
```

# Net growth
```{r}
lettuce_adj%>%
  select(Strain,Condition,Replicate,Day,cfu_lettuce) %>%
 filter(Strain!="APC") %>%
     filter(Strain!="MacConkey") %>%
    filter(!is.na(cfu_lettuce)) %>%
  spread(key=Day,value=cfu_lettuce,sep="_") %>%
  mutate(netgrowth=log10(Day_7/Day_Inoculum)) -> lettuce_adj2

lettuce_adj2 %>%  
  group_by(Strain, Condition) %>%
  summarize(average=mean((netgrowth)),
            SDE=sd((netgrowth))/sqrt(n()),
            n=n()) -> lettuceSummary_netgrowth

format.data.frame(lettuceSummary_netgrowth, dec=".") ->netgrowth_Ecoli
write_xlsx(netgrowth_Ecoli, "netgrowth_Ecoli.xlsx")
```