---
title: "Tomato_git"
author: "Sophia"
Date: 12.11.2017
output:
  html_document:
    code_folding: hide
    toc: true
---

```{r echo=F, message=F, warning=F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lme4)
library(lsmeans)
library(ggthemes)
library(emmeans)
library(writexl)
```

#Experimental Design 
 * Tomatoes were inoculated with 100uL bacterial culture
 * 3 tomatoes were placed in 1 bag
 * 30 mL of PBS was added into each bag for processing
 * Dilutions of wash water was plated

# Import and clean up the spreadsheet
 *All Variables need to be factor or numeric
 *Substitute detection limits, lower: 300CFU, upper: 4e5
 *Raw count in excel is CFU/mL from Qcount, taken into account dilutions but not 30 mL PBS
 
 
```{r}
tomato <- read.csv("Tomato_Summary_strain_FSL.csv")
#make sure all variables are factor or numeric if needed
tomato_adj <- tomato %>%
  mutate(Replicate=as.factor(Replicate), Dilution=as.numeric(Dilution))  %>%
  # Substitute upper detection limit for tntc counts
  mutate(Raw=ifelse(tnt.cnt.no_growth=="tntc",(4e5)*10^Dilution,Raw)) %>%
  # Substitute lower detection limit for zero counts
  mutate(Raw=ifelse(tnt.cnt.no_growth=="no_growth",10*10^Dilution,Raw)) %>%
  # Calculate inoculum per-tomato: Raw/10, since 100uL was added per tomato
  # Calculate wash water per-tomato: Raw*10
  mutate(cfu_tomato=ifelse(Day=="Inoculum",Raw/10,Raw*10)) %>%
  #Divide by 3 because we had 3 tomatoes 
  mutate(cfu_singletomato=cfu_tomato/3)


tomato_adj %>%  
  group_by(Strain, Condition, Day) %>%
  filter(!is.na(cfu_singletomato)) %>%
  summarize(average=mean(log10(cfu_singletomato)),
            SDE=sd(log10(cfu_singletomato))/sqrt(n()),
            n=n(),
            zeroes=sum(cfu_singletomato==0)) -> tomatoSummary
```

#Average inoculum
```{r}
tomato_adj %>%  
  filter(Day=="Inoculum")%>%
 # group_by(Strain) %>%
  summarize(average=mean(log10(cfu_singletomato)),
            SDE=sd(log10(cfu_singletomato))/sqrt(n()),
            n=n(),
            zeroes=sum(cfu_singletomato==0)) -> tomatoSummary_Inoculum
```


#Arithemtic mean for aerobic plate count (APC)
```{r}
#Each APC count
tomato_APC <- tomato_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_tomato) %>%
  filter(Strain=="APC") %>%
  spread(key=Day,value=cfu_tomato,sep="_") %>%
  mutate(APC_day7_day0=log10(Day_7/Day_0))
View(tomato_APC)

#Summary APC per day
tomato_adj %>%  
  filter(Strain=="APC")%>%
  group_by(Day) %>%
  summarize(average=mean(log10(cfu_singletomato)),
            SDE=sd(log10(cfu_singletomato))/sqrt(n()),
            n=n(),
            zeroes=sum(cfu_singletomato==0)) -> tomatoSummary_APC

#APC on day 7 compared to day 0 
tomato_APC %>%  
  summarize(average=mean(APC_day7_day0),
            SDE=sd(APC_day7_day0)/sqrt(n()),
            n=n()) -> tomato_APC_adj
View(tomato_APC_adj)

```

#Plots of raw counts
```{r}
#Plot Strain~Condition with each of the three replicates
tomato_adj %>%
  filter(Day != "Inoculum") %>%
  mutate(Day_n = as.numeric(as.character(Day))) %>%
  ggplot(aes(x=Day_n,y=cfu_singletomato,color=Replicate)) +
  geom_point() +
  geom_line() +
  scale_y_log10() +
  facet_grid(Strain~Condition)

#Plot with arithemtic mean for Strain~Condition and add fancy ribbon for standard deviation, write in bold processing days

#Write out full Food Safety Lab (FSL) strain ID
tomatoSummary$Strain2 <- factor(tomatoSummary$Strain, levels=c("FSL R9-5272", "FSL R9-5344", "FSL R9-5400", "FSL R9-5502", "FSL R9-6232", "APC")) #re-order how it is displayed
tomatoSummary$Day2 <- factor(tomatoSummary$Day, levels=c("Inoculum", "0", "1", "3", "4", "7")) #re-order how it is displayed
levels(tomatoSummary$Day2)[levels(tomatoSummary$Day2)=="Inoculum"] <- "Inoc"
levels(tomatoSummary$Strain2)[levels(tomatoSummary$Strain2)=="FSL R9-6232"] <- "FSL R9-6232*"
tomatoSummary %>%
  ggplot(aes(x=Day2,y=average, color=Strain2, group=1)) +
 guides(color=FALSE)+
    labs(x="Day",y="log CFU/tomato",title=expression(italic("Salmonella"))) +
  geom_point() +
    geom_ribbon(aes(ymin=average-SDE,
                  ymax=average+SDE,fill=Strain2, linetype=NA), alpha=0.2) + 
     scale_x_discrete(limits=c("Inoc","0","1","2","3","4","5", "6","7"))+
  theme(axis.text.x = element_text(face = c('bold', 'bold', 'bold', 'plain', 'bold', 'bold', 'plain', 'plain', 'bold'), size = c('10.8', '11', '11', '9', '11', '11', '9', '9', '11'), margin=margin(-35,0,0,0)),legend.position="none")+
 #labs( caption="Ribbon: Standard deviation")+
   labs( caption="*Surrogate organism")+
  facet_grid(Condition~Strain2) 
ggsave("Tomato.png", width=18, height=9,  dpi=600, units=c("in"))
```

#Population dynamics will be assessed for three phases (i) initial die-off, (ii) change in bacterial numbers over the initial 24 h, and (iii) growth or die-off over 7 day storage

# (i) Initial die-off
```{r}
tomato_die_off_sur <- tomato_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_singletomato) %>%
 filter(Strain!="APC") %>%
  spread(key=Day,value=cfu_singletomato,sep="_") %>%
  mutate(initial_die_off_sur=log10(Day_0/Day_Inoculum))

# Fit (fixed-effect) linear model
m_die_off_anova_sur <- lm(initial_die_off_sur~Strain*Condition,
                  data=tomato_die_off_sur)
anova(m_die_off_anova_sur)

#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_die_off_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1


# Since interaction is not significant, remove before doing pairwise tests:
m_die_off_anova <- update(m_die_off_anova_sur,.~.-Strain:Condition)
anova(m_die_off_anova_sur)
qqnorm(resid(m_die_off_anova_sur))

qqline(resid(m_die_off_anova_sur))
plot(predict(m_die_off_anova_sur),resid(m_die_off_anova_sur))

die_off_lsm_condition_sur <- lsmeans(m_die_off_anova_sur,~Condition)
die_off_pairwise_condition_sur <- lsmeans(die_off_lsm_condition_sur,pairwise~Condition)
condition_cld_sur <- cld(die_off_pairwise_condition_sur[[1]],Letters=LETTERS) %>% select(Condition,.group)
summary(die_off_lsm_condition_sur) %>%
  data.frame() %>%
  left_join(condition_cld_sur,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
    labs(x="Condition",y="estimated means log reduction",title=(expression("Initial die off "(italic("Salmonella"))))) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)
summary(die_off_lsm_condition_sur)


#and now by strain
die_off_lsm_strain_sur <- lsmeans(m_die_off_anova_sur,~Strain)
die_off_pairwise_strain_sur <- lsmeans(die_off_lsm_strain_sur,pairwise~Strain)

strain_cld_sur <- cld(die_off_pairwise_strain_sur[[1]],Letters=LETTERS) %>% select(Strain,.group)
summary(die_off_lsm_strain_sur) %>%
  data.frame() %>%
  left_join(strain_cld_sur,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/tomato with confidence intervals for supplemental table 1
summary(die_off_lsm_strain_sur)
summary(die_off_lsm_condition_sur)

```

# (ii) change in bacterial numbers over the initial 24 h
```{r}

tomato_24h_sur <- tomato_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_singletomato) %>%
 filter(Strain!="APC") %>%
  spread(key=Day,value=cfu_singletomato,sep="_") %>%
  mutate(first_24h_sur=log10(Day_1/Day_0))

# Fit (fixed-effect) linear model
m_24h_anova_sur <- lm(first_24h_sur~Strain*Condition,
                  data=tomato_24h_sur)
anova(m_24h_anova_sur)
#Add effect size to ANOVA output - partial eta-squared
anova1<-anova(m_24h_anova_sur)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1

# Since interaction is not significant, remove before doing pairwise tests:
m_24h_anova_sur <- update(m_24h_anova_sur,.~.-Strain:Condition)
anova(m_24h_anova_sur)
qqnorm(resid(m_24h_anova_sur))
qqline(resid(m_24h_anova_sur))
plot(predict(m_24h_anova_sur),resid(m_24h_anova_sur))


first_24h_lsm_condition_sur <- lsmeans(m_24h_anova_sur,~Condition)
first_24h_pairwise_condition_sur <- lsmeans(first_24h_lsm_condition_sur,pairwise~Condition)
condition_cld_24h_sur <- cld(first_24h_pairwise_condition_sur[[1]], Letters=LETTERS) %>% select(Condition,.group)
summary(first_24h_lsm_condition_sur) %>%
  data.frame() %>%
  left_join(condition_cld_24h_sur,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) +
   labs(x="Condition",y="estimated means log reduction",title=(expression("First 24 h  "(italic("Salmonella"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)


#and now by strain
first_24h_lsm_strain_sur <- lsmeans(m_24h_anova_sur,~Strain)
first_24h_pairwise_strain_sur <- lsmeans(first_24h_lsm_strain_sur,pairwise~Strain)
strain_cld_24h_sur <- cld(first_24h_pairwise_strain_sur[[1]],Letters=LETTERS) %>% select(Strain,.group)
summary(first_24h_lsm_strain_sur) %>%
  data.frame() %>%
  left_join(strain_cld_24h_sur,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/tomato with confidence intervals for supplemental table 1
summary(first_24h_lsm_strain_sur)
summary(first_24h_lsm_condition_sur)
```

(iii) growth over 7 day storage

```{r}
tomato_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_singletomato) %>%
  filter(Strain!="APC") %>%
  spread(key=Day,value=cfu_singletomato,sep="_") -> tomato_adj_day


#Identify lowest count throughout 7 days incubation period
tomato_Nmin <- transform(tomato_adj_day, Nmin=pmin(Day_0, Day_1, Day_3, Day_4, Day_7))
tomato_Nmin %>%
  mutate(compare_Day7_Nmin=log10(Day_7/Nmin)) -> tomato_N7_Nmin

tomato_N7_Nmin %>%
  group_by(Strain,Condition) %>%
  summarize(median=ifelse(n()==3,median(compare_Day7_Nmin),NA),
            max=max(compare_Day7_Nmin),
            min=min(compare_Day7_Nmin)) %>%
  ggplot(aes(x=Condition,y=median,color=Strain)) +
  geom_point(position = position_dodge(width=.5)) +
  labs(caption="Bars represent minimum and maximum values; point represents median") +
  geom_errorbar(aes(ymin=min,ymax=max),position = position_dodge(width = .5))


# Fit (fixed-effect) linear model
m_tomato_N7_Nmin_anova <- lm(compare_Day7_Nmin~Strain*Condition,
                  data=tomato_N7_Nmin)
anova(m_tomato_N7_Nmin_anova)
#Add effect size to ANOVA output - partial eta-squared - partial eta-squared
anova1<-anova(m_tomato_N7_Nmin_anova)
anova1$pes <- c(anova1$'Sum Sq'[-nrow(anova1)], NA)/(anova1$'Sum Sq' + anova1$'Sum Sq'[nrow(anova1)])
anova1



# Since interaction is not significant, remove before doing pairwise tests:
m_tomato_N7_Nmin_anova <- update(m_tomato_N7_Nmin_anova,.~.-Strain:Condition)
anova(m_tomato_N7_Nmin_anova)
qqnorm(resid(m_tomato_N7_Nmin_anova))
qqline(resid(m_tomato_N7_Nmin_anova))
plot(predict(m_tomato_N7_Nmin_anova),resid(m_tomato_N7_Nmin_anova))


N7_Nmin__condition <- lsmeans(m_tomato_N7_Nmin_anova,~Condition)
N7_Nmin__pairwise_condition <- lsmeans(N7_Nmin__condition,pairwise~Condition)
condition_cld_N7_Nmin <- cld(N7_Nmin__pairwise_condition[[1]],Letters=LETTERS) %>% select(Condition,.group)
summary(N7_Nmin__condition) %>%
  data.frame() %>%
  left_join(condition_cld_N7_Nmin,by="Condition") %>%
  ggplot(aes(x=Condition,y=lsmean)) +
  ggtitle("N7-Nmin",subtitle=NULL)+
  labs(x="Condition",y="estimated means log CFU/tomato",title=(expression("Regrowth until day 7  "(italic("Salmonella"))))) +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)


#and now by strain
N7_Nmin__strain <- lsmeans(m_tomato_N7_Nmin_anova,~Strain)
N7_Nmin__pairwise_strain <- lsmeans(N7_Nmin__strain,pairwise~Strain)
strain_cld_N7_Nmin <- cld(N7_Nmin__pairwise_strain[[1]],Letters=LETTERS) %>% select(Strain,.group)
summary(N7_Nmin__strain) %>%
  data.frame() %>%
  left_join(strain_cld_N7_Nmin,by="Strain") %>%
  ggplot(aes(x=Strain,y=lsmean)) +
  ggtitle("N7-Nmin",subtitle=NULL)+
    labs(x="Condition",y="lsmeans") +
  geom_point(position = position_dodge(width=.5)) +
  geom_errorbar(aes(ymin=lower.CL,ymax=upper.CL),
                position = position_dodge(.5),width=0.4) +
  geom_text(aes(label=.group,y=upper.CL+.01),vjust=0)

#Estimated means log CFU/tomato with confidence intervals for  supplemental table 1
summary(N7_Nmin__strain)
summary(N7_Nmin__condition)

```


#We want to determine Days for Daymin and Daymax

```{r}
tomato_adj %>%
  select(Strain,Condition,Replicate,Day,cfu_tomato) %>%
  filter(Strain!="APC") %>%
  filter(Day != "Inoculum") %>%
  filter(Day != "0") %>%
  spread(key=Day,value=cfu_tomato,sep="_") -> tomato_adj_Days

#----min values
tomato_adj_Days_Nmin <- transform(tomato_adj_Days, Nmin=pmin(Day_1, Day_3, Day_4, Day_7))

tomato_adj_Days_Nmin->a_min

a_min$CompareColumn <- apply(a_min[, 4:7] == a_min[, 8], 1, function(x) ifelse(any(x), paste(colnames(a_min)[4:7][x], collapse=', '), 'None'))

count(a_min, c("CompareColumn")) ->b_min

#----- max value
tomato_adj_Days_Nmax <- transform(tomato_adj_Days, Nmax=pmax(Day_1, Day_3, Day_4, Day_7))

tomato_adj_Days_Nmax->a_max

a_max$CompareColumn <- apply(a_max[, 4:7] == a_max[, 8], 1, function(x) ifelse(any(x), paste(colnames(a_max)[4:7][x], collapse=', '), 'None'))

count(a_max, c("CompareColumn")) ->b_max
```


# Net growth
```{r}
tomato_adj%>%
  select(Strain,Condition,Replicate,Day,cfu_singletomato) %>%
 filter(Strain!="APC") %>%
  spread(key=Day,value=cfu_singletomato,sep="_") %>%
  mutate(netgrowth=log10(Day_7/Day_Inoculum)) -> tomato_adj2

tomato_adj2 %>%  
  group_by(Strain, Condition) %>%
  summarize(average=mean((netgrowth)),
            SDE=sd((netgrowth))/sqrt(n()),
            n=n()) -> tomatoSummary_netgrowth

format.data.frame(tomatoSummary_netgrowth, dec=".") ->netgrowth_Salmonella
write_xlsx(netgrowth_Salmonella, "netgrowth_Salmonella.xlsx")

#Arithemtic mean of bacteria recovered on day 7
tomato_adj2 %>% 
  group_by(Condition) %>% 
  summarize(average=log10(mean((Day_7))),
            SDE=log10(sd((Day_7)))/sqrt(n()),
            n=n()) -> Day_7_mean
  
#High salt per day
tomato_adj%>%
  select(Strain,Condition,Replicate,Day,cfu_singletomato) %>%
 filter(Strain!="APC") %>%
  filter(Condition=="NaCl") %>%
  group_by(Day) %>% 
  summarize(average=log10(mean((cfu_singletomato))),
            SDE=log10(sd((cfu_singletomato)))/sqrt(n()),
            n=n()) -> NaCl_mean
  

```

